 

### 2022-05-17 thesis/introduction.qmd
Adjust to new citation style

Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion. In a laboratory setting such tests commonly take place on treadmills or cycling ergometer. Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value. Therefore most exercise protocols consist of a continuous or stepwise increase with an optimal during around 8-12 minutes [@buchfuhrer1983 ; @yoon2007]. While this narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008]. Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.

Early research on V̇O~2max~ used bags to collect expired air for later analysis--- the so called Douglas bag method [@shephard2017]. Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001]. While mixing chambers are regarded as more reliable [@beijst2013], measuring breath-by-breath generates data with a higher temporal resolution [@roecker2005]. The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently.

@@

Different data processing strategies influence measured parameters of gas exchange. As soon as when the first automated systems for gas exchange measurement were available, @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values. Since then, many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that shorter calculation intervals lead to higher V̇O~2max~ values [@johnson1998; @astorino2009; @sell2021]. @midgley2007 found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes. Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

@@

Differences in V̇O~2max~ due to data processing can cause serious implications in practice. They hinder the comparability of data from studies within meta-analysis [@martin-rincon2019] or the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion of V̇O~2max~ [@astorino2009]. Crucially, in situation where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data, @robergs2003 opposed these conclusion, stating that shorter calculation intervals are less erroneous. They further argued to use breath-based moving averages instead of time averages. In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003]. In a classic opinion piece, @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by @weir2004. However, @robergs2010 acknowledged the lack of accessible software implementing such procedures.

In absence of one commonly accepted method, data processing varies among the literature. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. While providing interesting first insights into reporting and processing practices, the search by @midgley2007 was not systematic and its methods were not described in detail. @robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They provided the latter one with a total of 75 respondents, which reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning. Practices in reporting and processing may have changed more than a decade after the publication of such numbers.

@@

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. @martin-rincon2019 provided formula for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts. Therefore motivation and measurement devices may have interacted with the influence of processing strategies.

>>>>

Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion. In a laboratory setting such tests commonly take place on treadmills or cycling ergometer. Fatigue prior to reaching the V̇O~2max~ results in underestimating its true value. Therefore most exercise protocols consist of a continuous or stepwise increase in load with an optimal duration of 8-12 minutes [@buchfuhrer1983 ; @yoon2007]. While these narrow limits have been challenged, considerably shorter or longer protocols should not be used [@midgley2008]. Despite protocol type and protocol duration, factors such as motivation [@midgley2018], exercise modality [@myers1991] and biological variability [@katch1982] bias the V̇O~2max~ determination.

Early research on V̇O~2max~ used bags to collect expired air for later analysis --- the so called Douglas bag method [@shephard2017]. Today's modern metabolic carts allow for simultaneous collection and analysis of gases, using either mixing chambers (which often sample data at fixed time intervals, e.g., 15 seconds) or breath-by-breath methods [@macfarlane2001]. While mixing chambers are regarded as more reliable [@beijst2013], measuring breath-by-breath generates data with a higher temporal resolution [@roecker2005]. The measuring method and the metabolic cart model can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. But even the same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently.

@@

Different data processing strategies influence measured parameters of gas exchange. As soon as when the first automated systems for gas exchange measurement were available, Matthews et al. @matthews1987 reported that different processing strategies lead to different V̇O~2max~ values. Since then, many researchers investigated the influence of different data averaging intervals on V̇O~2max~, unsurprisingly showing that shorter calculation intervals lead to higher V̇O~2max~ values [@johnson1998; @astorino2009; @sell2021]. @midgley2007 Midgley et al. found differences between data processing strategies for V̇O~2max~, but no difference in the reliability of these, infering that no optimal strategy exists, but that consistency of strategies is key when comparing outcomes. Based on own data in sedentary and moderately trained individuals, @martin-rincon2019 Martin-Rincon et al. provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. There is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

@@

Differences in V̇O~2max~ due to data processing can cause serious implications in practice. They hinder the assessment of longitudinal data from athletes that participated in diagnostics with differing data processing or the comparability of data from studies within meta-analysis [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009]. Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013a; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

Calls to standardize data processing strategies of gas exchange data are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. Howley et al. @howley1995 argued to use longer calculation intervals (60 seconds), as shorter intervals may introduce bias towards extreme data values and thus could systematically overestimate true V̇O~2max~. Based on synthetic and experimental data, Robergs & Burnett @robergs2003 opposed this conclusion, stating that shorter calculation intervals are less erroneous. They further argued to use breath-based moving averages instead of time averages. In terms of data processing, the current ATS/ACCP guidelines for cardiopulmonary exercise testing remain vague, stating that *"30 to 60-second intervals for averaging data are recommended, although 20-seconds intervals may be acceptable*" [@ats/accp2003]. In a classic opinion piece, Robergs et al. @robergs2010 argued for using digital filtering (i.e. a low-pass Butterworth filter) for processing gas exchange data, a strategy initially introduced for gas exchange measures by Weir et al. @weir2004. However, @robergs2010 Robergs et al. acknowledged the lack of accessible software implementing such procedures.

In absence of one commonly accepted method, data processing varies among the literature. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research articles [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. @midgley2007 Midgley et al. were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting their methods choose binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. While providing interesting first insights into reporting and processing practices, the search by Midgley et al. @midgley2007 was not systematic and its methods were not described in detail. Robergs et al. @robergs2010 wrote that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They provided the latter one with a total of 75 respondents, who reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Astonishingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective influences as opposed to objective reasoning. Practices in reporting and processing may have changed more than a decade after the publication of such numbers.

@@

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these works lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. @martin-rincon2019 Martin-Rincon et al. provided formulas for comparing data processing strategies by investigating a data set of sedentary or recreational trained athletes, using two different metabolic carts. Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies.

### 2022-05-17 thesis/results.qmd
Adjust to new citation style

Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in @fig-flow). Out of the random sample (n = 500), 242 articles were included in the final analysis.

@@

For preprocessing, some authors reported the use of a (linear) interpolation for the breath-by-breath data to seconds (n = 7; 4.3%). A minority of researchers reported the use of data filtering strategies to remove outliers. This included the use of initial data smoothing by a short moving average (3 seconds, n = 1; 5 breaths, n = 3), the manual detection and removal of outliers (n = 2) or an automated removal of outliers (n = 5). For the automated outlier detection authors removed single data points differing from a local mean by a varying number of standard deviations (2, 3 or 4) or being outside of a 95% confidence interval. When reported, the software used for data processing varied among studies showing a total of more than 15 reported programs (for 30 studies that reported this parameter).

The calculation interval for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

@@

Binned time averages systematically generate V̇O~2max~ values compared to moving averages. Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values. Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.

>>>>

Initial search yielded 7529 results of which 4364 remained after automated filtering and removal of duplicates (see flow diagram in @fig-flow). Out of the random sample (n = 500), 242 articles were included in the final analysis. All sampled studies and their inclusion/exclusion status and reason are available on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/fullsample.csv).

@@

For preprocessing, some authors reported the use of a (linear) interpolation for the breath-by-breath data to seconds (n = 7; 4.3%). A minority of researchers reported the use of data filtering strategies to remove outliers. This included the use of initial data smoothing by a short moving average (3 seconds, n = 1; 5 breaths, n = 3), the manual detection and removal of outliers (n = 2) or an automated removal of outliers (n = 5). For the automated outlier detection authors removed single data points differing from a not further defined local mean by a varying number of standard deviations (2, 3 or 4) or being outside of a 95% confidence interval. When reported, the software used for data processing varied among studies showing a total of more than 15 reported programs (for 30 studies that reported this parameter).

The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

@@

Binned time averages systematically generate lower V̇O~2max~ values compared to moving averages. Decreasing the calculation interval to 5 seconds leads to a 3-4% median increase of V̇O~2max~ values. Notably on an individual level these increases may be lower (≈2%) or much higher (\>10%) than the median value.
